\documentclass[letterpaper,twocolumn,10pt]{article}

\usepackage{usenix-2020-09}

% ---- Packages ----
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{multirow}

% ---- Shared notation and theorem environments ----
\input{math_commands}

\begin{document}

% ---- Title ----
\title{SafeMimic: Dependency-Aware Adversarial Mimicry Attacks\\
on Provenance Graph Intrusion Detection}

\author{{\rm Anonymous Submission}\\
Paper \#XXX}

\maketitle

% ---- Abstract ----
\begin{abstract}
Provenance graph-based host intrusion detection systems (HIDS) are widely
deployed to detect advanced persistent threats (APTs). Recent mimicry attacks
attempt to evade these detectors by inserting benign-looking operations into
attack workflows. However, existing approaches face a fundamental dual failure:
inserted operations may break attack functionality (the safety problem), and the
evasion process itself creates detectable artifacts such as excess edges and
skewed frequency distributions (the stealthiness problem). We present
SafeMimic, a dependency-aware adversarial mimicry framework that formally
addresses both problems simultaneously. SafeMimic first extracts the attack's
resource dependency set and each candidate operation's effect set, constructing
a provably safe pool of operations that cannot interfere with the attack. It
then selects from this safe pool using KL-divergence minimization and temporal
smoothing to match the target environment's normal behavior distribution. We
prove that SafeMimic guarantees attack preservation (Safety Theorem) and show
that the safe pool expands dynamically as the attack progresses. Evaluation on
four APT attack scenarios from the DARPA Transparent Computing program demonstrates 100\% attack preservation, while
experiments on three DARPA TC datasets (CADETS, THEIA, TRACE) show that SafeMimic
achieves 0\% ML detection rate---a 95\% reduction in anomaly scores.
Surprisingly, we find that ProvNinja's regularity-based camouflage
\emph{backfires}, increasing ML detection from 94\% to 100\% by creating its
own anomalous behavioral pattern.
\end{abstract}

% ---- Sections ----
\input{sections/introduction}
\input{sections/background}
\input{sections/design}
\input{sections/implementation}
\input{sections/evaluation}
\input{sections/related}
\input{sections/discussion}
\input{sections/conclusion}

% ---- Bibliography ----
\bibliographystyle{plain}
\bibliography{references}

\end{document}
